Towards interactive model checking\\of biological pathway dynamics

### 2015-01-22 Paper_FORTE15.tex
Latest corrections and new changes to make it fit 15 pages

\author{Stefano Schivo${}^\star$
\and Jetse Scholma${}^\star$
\and Paul E. van der Vet
\and Marcel Karperien
\and Janine N. Post
\and Jaco van de Pol${}^{\star \star}$
\and Rom Langerak${}^{\star \star}$}

@@

The current knowledge on signaling pathways (mostly organized in databases such as KEGG~\cite{kegg}
or PhosphoSite~\cite{phosphosite}) suggests that the interactions involved in a cellular response assume more often
the shape of a network than that of a simple chain of signal relays.
Such networks are typically highly connected, involving feedback loops and crosstalk
between multiple pathways, making it difficult for the human
brain to grasp their dynamic behavior.
For this reason, a computational support is required when studying non-trivial biological networks.

@@

The current knowledge on signaling pathways~\cite{kegg,phosphosite} evidences the fact
that signaling interactions are rarely a simple chain of activations as represented in Figure~\ref{fig:small-example-net}.
More often, they assume the shape of a network with multiple feedback loops and crosstalk from different signaling sources.
This complexity is an added difficulty for the study of such networks, reducing the possibilities
%for the human brain alone
to deduce the dynamic behavior of a signaling network by inspecting its static representation.
For this reason, an efficient computational support is essential when representing and analyzing the behavior
of complex signaling networks.

@@

The occurrence of a \emph{reaction} modifies by one discrete step the activity level of its target reactant, making it increase or decrease 
depending on whether the reaction is defined, respectively, as activating or inhibiting.
The rate with which a reaction occurs depends on a formula selected by the user. 
Choosing one of three available scenarios allows the user to make the reaction rate depend on
the activity of one or two reactants.
The rate of each reaction can be scaled by modifying the value of one kinetic constant $k$,
possibly using a qualitative measure from a predefined set ({\sf very slow}, {\sf slow},
{\sf medium}, {\sf fast}, {\sf very fast}).
% Approximating the actual reaction dynamics with less detailed single-parameter scenarios
% leads to a kinetic model that is still precise enough to faithfully represent the behavior of a signaling network.
The approximation allows us to reduce the dependence of a model from often unavailable quantitative
parameters for biochemical reaction kinetics, while keeping a precision level that is still precise enough to be useful.
For a more precise dissertation on how reaction rates are computed in ANIMO, we recommend~\cite{animo-ieee},
where the previously used \tas\ model is presented. The reader interested in the current
methods for parameter setting in ANIMO can refer to~\cite{animo-syncop}.

@@

\begin{figure}[thb]
  \begin{center}
  \subfloat[\label{fig:network-abcd}]{\begin{minipage}[c][\reactantTAheight]{0.26\textwidth}\begin{center}\includegraphics[scale=1.0]{images/network_abcd}\end{center}\end{minipage}} \qquad\qquad\quad
  \subfloat[$R_A$\label{fig:ta-new-model}]{\begin{minipage}[c][\reactantTAheight]{0.6\textwidth}\begin{center}\reactantTA\end{center}\end{minipage}} \\
  \subfloat[$R_1: B \rightarrow A$ \label{fig:ta-old-BA}]{\includegraphics[scale=\scalaOldModel]{images/Reaction_B_A}} \qquad\qquad
%   \subfloat[$R_2: C \rightarrow A$ \label{fig:ta-old-CA}]{\includegraphics[scale=\scalaOldModel]{images/Reaction_C_A}} \qquad\qquad
  \subfloat[$R_3: D \dashv A$ \label{fig:ta-old-DA}]{\includegraphics[scale=\scalaOldModel]{images/Reaction_D_A}}
  \end{center}\vspace{-.5cm}
  \caption{
{\bf\protect\subref{fig:network-abcd}} A signaling network where one node is influenced by three distinct reactions.
The (virtual) reaction $R_A$ is defined in
the reactant-centered model as the algebraic sum of the reactions influencing $A$.
({\bf\protect\subref*{fig:ta-new-model}\--{}\protect\subref*{fig:ta-old-DA}}) \tas\ templates generated by ANIMO for the reactant-centered~{\bf\protect\subref{fig:ta-new-model}}
 and reaction-centered~({\bf\protect\subref*{fig:ta-old-BA}\--{}\protect\subref*{fig:ta-old-DA}}) models of the network in~{\bf\protect\subref{fig:network-abcd}}.
The templates have been edited for an easier visual comparison; while the template for $R_2$ ($C \rightarrow A$) is qualitatively identical to the one for $R_1$,
and was left out for the sake of space. Functions {\sf can\_react()} and {\sf cant\_react()}
in~{\protect\subref{fig:ta-new-model}} have the same content as the corresponding guards in ({\bf\protect\subref*{fig:ta-old-BA}\--{}\protect\subref*{fig:ta-old-DA}}),
while function {\sf update()} in~{\protect\subref{fig:ta-new-model}}
replaces the references to the pre-computed tables {\sf tL[]}, {\sf tU[]} with on-the-fly computation (see Sect.~\ref{sec:rates-ta-model}).
% \ta\ template used to represent a net reaction in the reactant-centered model.
\label{fig:ta-models}}
\end{figure}

As with ODEs, if $r_A$ is positive the activity level of $A$ will increase;
otherwise, $A$ will decrease its activity level.
The absolute value of $r_A$ determines the speed with which such change happens.
The value of the reaction rate is thus translated into a time value to be used as
time bound in the \ta\ representing $R_A$ (see Fig.~\ref{fig:ta-new-model}) by computing
$$T_A = \frac{1}{\mbox{abs}(r_a)}$$
with $\mbox{abs}(r_a)$ the absolute value of $r_a$. In order to represent
a natural variability in reaction timing, we relax the exact value of $T_A$ by
defining bounds of~$\pm~5\%$ (which can be changed by the user):
$$
\begin{array}{lcl}
  \mbox{\sf tL} &=& T_A \times 0.95 \\
  \mbox{\sf tU} &=& T_A \times 1.05
\end{array}
$$
Analogously to the reaction-centered model of~\cite{animo-ieee},
we will call {\sf tL} the \emph{lower time bound} and {\sf tU} the \emph{upper time bound}.
Note that here, differently from what presented in~\cite{animo-ieee}, we do not define
pre-computed tables with all possible values for {\sf tL} and {\sf tU}: this is because
the number of variables for the computation of a reaction rate can be arbitrarily high, and
the size of such tables grows exponentially with the number of variables.
In the example of Figure~\ref{fig:ta-models}, a table to represent the time bounds
for reaction ${R_A}$ would have four dimensions, depending on the activity levels
of reactants $A$, $B$, $C$ and $D$. Assuming that all four reactants are discretized
on 101 levels (activity levels from 0 to 100), each of the two tables {\sf tL} and {\sf tU} would contain $101^4 = 104\,060\,401$
elements. Adding one more reaction $E \rightarrow A$ to the pool would make
this approach exceed the memory limits of most of the currently available computers
% , requiring about 78 gigabytes of memory to be stored.
For this reason, the reaction rates are computed on the fly (see Section~\ref{sec:rates-ta-model}).

@@

Being based on the same abstractions, the two models can be analyzed with the same
UPPAAL queries for all the cases for which biology is involved. The differences can be seen only if the
analysis focuses on the single automaton level and explicitly refers to locations: such type of analysis
is normally not necessary when considering the model as a whole and using it for what it is, i.e. an
abstract representation of a set of biological processes.

@@

We note that performing model checking means dealing with state space explosion problems,
and model reduction is recommendable in order to obtain any result within adequate time limits. One of the
most user-accessible ways available in ANIMO to reduce the size of a model is setting the uncertainty to 0
before performing model checking queries.
In order to still consider some biological variability, the user can manually perform multiple model checking queries with changed interaction parameters.
The tests performed in this section have an uncertainty level set to 0, instead of the $5\%$ recommended for simulation-based experiments, to
make model checking feasible within seconds.

@@

\section{Conclusions and future work}\label{sec:conclusion}
We have presented here how the ANIMO tool 
was improved to provide a more interactive modeling process.
Thanks to the increased performances of the new reactant-centered modeling approach,
we are able to obtain answers to model checking queries in a matter of seconds.
Coherently with the user friendliness of the tool, the features
of model checking are made accessible without the need to directly deal with \tas\ models.
In this way, ANIMO acts as an intermediary between the biologist and a formal
representation of biological signaling pathways, letting the experts concentrate
on investigating the mechanisms of cellular responses.

@@

We aim at widening the available set of model checking queries, in order to allow biologists to perform
in silico experiments on an already fitting model and to obtain answers to more relevant questions.
This would increase the usefulness of a model as a help to drive wet-lab investigation.
In order to allow for meaningful in silico experiments, we plan to purposefully introduce user-defined non-deterministic 
parts in our models, which would allow for drug dosage investigations through model checking.
This can be done e.g. through the definition of intervals for the values of some reaction kinetic constants,
adding considerable uncertainty in the timing of those reactions.
A model of this type could then be interrogated with queries like ``How much X, Y and Z do we need to provide,
and at which time points, in order to obtain targets A, B and C to be active?''.
More useful results can be achieved through proper application of the statistical model checking part of UPPAAL~\cite{uppaal-smc} to
an extended version of our model to include realistically significant stochastic behavior.
Finally, in order to further improve performances, the extension of ANIMO with support for a multi-core model checking approach based on the
works by Dalsgaard et al.~\cite{uppaal-multi-core1}
is under study.

>>>>

\author{Stefano~Schivo${}^\star$
\and Jetse~Scholma${}^\star$
% \and Paul E. van der Vet
\and Marcel~Karperien
\and Janine~N.~Post
\and Jaco~van~de~Pol${}^{\star \star}$
\and Rom~Langerak${}^{\star \star}$}

@@

The current knowledge on signaling pathways (mostly organized in databases such as KEGG~\cite{kegg}
or PhosphoSite~\cite{phosphosite}) suggests that the interactions involved in a cellular response assume more often
the shape of a network than that of a simple chain of signal relays.
Such networks are typically highly connected, involving feedback loops and crosstalk
between multiple pathways, making it difficult 
to grasp their dynamic behavior.
For this reason, computational support is required when studying non-trivial biological networks.

@@

The current knowledge on signaling pathways~\cite{kegg,phosphosite} evidences the fact
that signaling interactions are rarely a simple chain of activations as represented in Figure~\ref{fig:small-example-net}.
More often, they assume the shape of a network with multiple feedback loops and crosstalk from different signaling sources.
This complexity is an added difficulty for the study of such networks, reducing the possibilities
%for the human brain alone
to deduce the dynamic behavior of a signaling network by inspecting its static representation.
For this reason, efficient computational support is essential when representing and analyzing the behavior
of complex signaling networks.

@@

The occurrence of a \emph{reaction} modifies by one discrete step the activity level of its target reactant, making it increase or decrease 
depending on whether the reaction is defined, respectively, as activating or inhibiting.
The rate with which a reaction occurs depends on a formula selected by the user. 
Choosing one of three available scenarios allows the user to make the reaction rate depend on
the activity of one or two reactants.
The rate of each reaction can be scaled by modifying the value of one kinetic constant $k$,
possibly using a qualitative measure from a predefined set ({\sf very slow}, {\sf slow},
{\sf medium}, {\sf fast}, {\sf very fast}).
% Approximating the actual reaction dynamics with less detailed single-parameter scenarios
% leads to a kinetic model that is still precise enough to faithfully represent the behavior of a signaling network.
The approximation allows us to reduce the dependence of a model from often unavailable quantitative
parameters for biochemical reaction kinetics, while keeping a precision level that is still precise enough to be useful.
For a more precise explanation on how reaction rates are computed in ANIMO, we recommend~\cite{animo-ieee},
where the previously used \tas\ model is presented. The reader interested in the current
methods for parameter setting in ANIMO can refer to~\cite{animo-syncop}.

@@

\begin{figure}[thb]
  \begin{center}
  \subfloat[\label{fig:network-abcd}]{\begin{minipage}[c][\reactantTAheight]{0.26\textwidth}\begin{center}\includegraphics[scale=1.0]{images/network_abcd}\end{center}\end{minipage}} \qquad\qquad\quad
  \subfloat[$R_A$\label{fig:ta-new-model}]{\begin{minipage}[c][\reactantTAheight]{0.6\textwidth}\begin{center}\reactantTA\end{center}\end{minipage}} \\
  \subfloat[$R_1: B \rightarrow A$ \label{fig:ta-old-BA}]{\includegraphics[scale=\scalaOldModel]{images/Reaction_B_A}} \qquad\qquad
%   \subfloat[$R_2: C \rightarrow A$ \label{fig:ta-old-CA}]{\includegraphics[scale=\scalaOldModel]{images/Reaction_C_A}} \qquad\qquad
  \subfloat[$R_3: D \dashv A$ \label{fig:ta-old-DA}]{\includegraphics[scale=\scalaOldModel]{images/Reaction_D_A}}
  \end{center}\vspace{-.5cm}
  \caption{
{\bf\protect\subref{fig:network-abcd}} A signaling network where one node is influenced by three distinct reactions.
The (virtual) reaction $R_A$ is defined in
the reactant-centered model as the algebraic sum of the reactions influencing $A$.
({\bf\protect\subref*{fig:ta-new-model}\--{}\protect\subref*{fig:ta-old-DA}}) \tas\ templates generated by ANIMO for the reactant-centered~{\bf\protect\subref{fig:ta-new-model}}
 and reaction-centered~({\bf\protect\subref*{fig:ta-old-BA}\--{}\protect\subref*{fig:ta-old-DA}}) models of the network in~{\bf\protect\subref{fig:network-abcd}}.
The templates have been edited for an easier visual comparison; while the template for $R_2$ ($C \rightarrow A$) is qualitatively identical to the one for $R_1$,
and was left out for the sake of space. Functions {\sf can\_react()} and {\sf cant\_react()}
in~{\protect\subref{fig:ta-new-model}} have the same content as the corresponding guards in ({\bf\protect\subref*{fig:ta-old-BA}\--{}\protect\subref*{fig:ta-old-DA}}),
while function {\sf update()} in~{\protect\subref{fig:ta-new-model}}
replaces the references to the pre-computed tables {\sf tL[]}, {\sf tU[]} with on-the-fly computation (see Sect.~\ref{sec:rates-ta-model}).
% \ta\ template used to represent a net reaction in the reactant-centered model.
\label{fig:ta-models}}
\vspace{-1mm}
\end{figure}

As with ODEs, if $r_A$ is positive the activity level of $A$ will increase;
otherwise, $A$ will decrease its activity level.
The absolute value of $r_A$ determines the speed with which such change happens.
The value of the reaction rate is thus translated into a time value to be used as
time bound in the \ta\ representing $R_A$ (see Fig.~\ref{fig:ta-new-model}) by computing
$$T_A = \frac{1}{\mbox{abs}(r_a)}$$
with $\mbox{abs}(r_a)$ the absolute value of $r_a$. In order to represent
a natural uncertainty or variability in reaction timing, we relax the exact value of $T_A$ by
defining bounds of~$\pm~5\%$ (which can be changed by the user):
$$
\begin{array}{lcllcl}
  \mbox{\sf tL} &=& T_A \times 0.95 \qquad
  \mbox{\sf tU} &=& T_A \times 1.05
\end{array}
$$
Analogously to the reaction-centered model of~\cite{animo-ieee},
we will call {\sf tL} the \emph{lower time bound} and {\sf tU} the \emph{upper time bound}.
So we replace an exact reaction time by an interval of possible reaction times,
%  (this means that we
% implicitly assume a uniform distribution of the reaction time over this interval).
implicitly assuming a uniform distribution over this interval.

Differently from what presented in~\cite{animo-ieee}, we do not define
pre-computed tables with all possible values for {\sf tL} and {\sf tU}: this is because
the number of variables for the computation of a reaction rate can be arbitrarily high, and
the size of such tables grows exponentially with the number of variables.
In the example of Figure~\ref{fig:ta-models}, a table to represent the time bounds
for reaction ${R_A}$ would have four dimensions, depending on the activity levels
of reactants $A$, $B$, $C$ and $D$. Assuming that all four reactants are discretized
on 101 levels (activity levels from 0 to 100), each of the two tables {\sf tL} and {\sf tU} would contain $101^4 = 104\,060\,401$
elements. Adding one more reaction $E \rightarrow A$ to the pool would make
this approach exceed the memory limits of most of the currently available computers
% , requiring about 78 gigabytes of memory to be stored.
For this reason, the reaction rates are computed on the fly (see Section~\ref{sec:rates-ta-model}).

@@

Being based on the same abstractions, the two models can be analyzed with the same
UPPAAL queries as long as no explicit reference is made to \tas\ locations. Such an analysis
would normally not be necessary if considering the model as a whole and using it for what it is, i.e. an
abstract representation of a set of biological processes.

@@

We note that performing model checking means dealing with state space explosion problems,
and model reduction is recommendable in order to obtain any result within adequate time limits. One of the
most user-accessible ways available in ANIMO to reduce the size of a model is setting the ``natural uncertainty''
defined in Sect.~\ref{sec:reactant-centered} to 0 before performing model checking queries.
In order to still consider some biological variability, the user can manually perform multiple model checking queries with changed interaction parameters.
The tests performed in this section have an uncertainty level set to 0, instead of the $5\%$ recommended for simulation-based experiments, to
make model checking feasible within seconds.

@@

\section{Conclusions and future work}\label{sec:conclusion}
We have presented here how the ANIMO tool 
was improved to provide a more interactive modeling process.
Thanks to the increased performances of the new reactant-centered modeling approach,
we are able to obtain answers to model checking queries in a matter of seconds.
The features
of model checking are made accessible without the need to directly deal with \tas\ models.
In this way, ANIMO acts as an intermediary between the biologist and a formal
representation of biological signaling pathways, letting the experts concentrate
on investigating the mechanisms of cellular responses.

@@

We aim at widening the available set of model checking queries, in order to allow biologists to perform
in silico experiments on an already fitting model and to obtain answers to more relevant questions.
This would increase the usefulness of a model as a help to drive wet-lab investigation.
In order to allow for meaningful in silico experiments, we plan to purposefully introduce user-defined non-deterministic 
parts in our models, which would allow for drug dosage investigations through model checking.
This can be done e.g. through the definition of intervals for the values of some reaction kinetic constants,
adding considerable uncertainty in the timing of those reactions.
% A model of this type could then be interrogated with queries like ``How much X, Y and Z do we need to provide,
% and at which time points, in order to obtain targets A, B and C to be active?''.
More useful results can be achieved through proper application of the statistical model checking part of UPPAAL~\cite{uppaal-smc} to
an extended version of our model to include realistically significant stochastic behavior.
Finally, in order to further improve performances, the extension of ANIMO with support for a multi-core model checking approach based on the
works by Dalsgaard et al.~\cite{uppaal-multi-core1}
is under study.
